Variety in the Implementation of Nonlinear Least Squares Program Codes

### 2022-01-22 VarietyInNLS/VarietyInNonlinearLeastSquaresCodes.Rmd
Tidying article

- programming language

@@

# Saving storage

@@

For example, we can form sums of squares in R using `for` loops, a vectorized
`sum()` or the `crossprod()` function. These have different performance profiles.
We can time the sums of squares calculation of different length random vectors
using these three approaches and get the ratio to the time using `crossprod()`.
This reveals that the time for summing 100 elements is much longer than that
of summing 1000, which seems very strange. If we first "compile" the functions
using `compiler::cmpfun()` we get the second line of timings, which are ordered
by the size of the vector. Thus we are suspicious that R is trying to compile
the functions "on the fly" and this may be taking time when the program
is initiated. Timings in microseconds were run on a machine with the profile (@RMachID2022)

@@

```
 n         t(forloop) : ratio         t(sum) : ratio        t(crossprod)  all.equal 
100       26686.85 :  1.883607     12721.49  :  0.8979062   14167.95      TRUE 
           3968.35 :  2.074802       953.36  :  0.4984524    1912.64 
1000      33413.23 :  12.08929      2372.34  :  0.8583399    2763.87      TRUE 
          32894.62 :  11.96442      2319.56  :  0.8436696    2749.37 
10000    325475.7  :  28.14976     15694.36  :  1.357375    11562.29      TRUE 
         331670.1  :  28.24715     16511.91  :  1.40626     11741.72 
1e+05   3234528    :  31.81066    346601.8   :  3.408729   101680.6       TRUE 
        3178637    :  30.75105    346969     :  3.356677   103366.8 
1e+06  31728237    :  29.08711   4057012     :  3.719297  1090801         TRUE 
       31837624    :  29.11575   4051945     :  3.705535  1093485 
```

@@

# Implementation choices for an R package

??2021 update to nlsr and ideas for the "Improvements to nls()" project.

>>>>

- programming language (which might include called functions that have been
  constructed outside the current computing environment)

@@

## Saving storage

@@

For example, we can form sums of squares in R using `for` loops, a vectorized
`sum()` or the `crossprod()` function. These have different performance profiles.
We can time the sums of squares calculation of different length random vectors
using these three approaches and get the ratio to the time using `crossprod()`.
This reveals that the time for summing 100 elements may be much longer than that
of summing 1000, which seems very strange. If we first "compile" the functions
using `compiler::cmpfun()` we get the second line of timings, which are ordered
by the size of the vector. The third line times just the direct statements, and 
we see that explicit loops executed directly are quite inefficient unless compiled
first. Note that `crossprod()` calls code from whatever BLAS is linked to R (@BLAS79 
and subsequent evolution of these ideas).

Because of the differences and orderings of the timings, we are suspicious that R 
is trying to compile functions "on the fly" and this may be taking time when the program
is initiated. R is constantly being reworked, with new releases several times per
year, with possible changes in how program code is parsed and executed. 
Timings in microseconds were run on a machine with the profile (see @RMachID2022)

@@

```
 n         t(forloop) : ratio         t(sum) : ratio        t(crossprod)  all.equal 
100       26686.85 :  1.883607     12721.49  :  0.8979062   14167.95      TRUE 
           3968.35 :  2.074802       953.36  :  0.4984524    1912.64 
        1767952    :  1172.25       1120.33  :  0.7428407 	 1508.17 
1000      33413.23 :  12.08929      2372.34  :  0.8583399    2763.87      TRUE 
          32894.62 :  11.96442      2319.56  :  0.8436696    2749.37 
        1827674    :  747.1422 	    2645.11  :  1.081305 	   2446.22 
10000    325475.7  :  28.14976     15694.36  :  1.357375    11562.29      TRUE 
         331670.1  :  28.24715     16511.91  :  1.40626     11741.72 
        2090288    :  180.5873     15792.13  :  1.364338    11574.94 
1e+05   3234528    :  31.81066    346601.8   :  3.408729   101680.6       TRUE 
        3178637    :  30.75105    346969     :  3.356677   103366.8 
        5104998    :  48.30487    361242.5   :  3.418175 	 105682.9 
1e+06  31728237    :  29.08711   4057012     :  3.719297  1090801         TRUE 
       31837624    :  29.11575   4051945     :  3.705535  1093485 
       33884253    :  27.14376 	 4560980     :  3.653678  1248326 
```

@@

<!-- # Implementation choices for an R package -->

<!-- ??2021 update to nlsr and ideas for the "Improvements to nls()" project. -->
